High-order cognition is supported by complex but compressible brain activity patterns

### 2023-02-28 paper/main.tex
typo fix in methods, added keywords, working on results section

\documentclass[english]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\usepackage{soul}
\linenumbers

@@

@@

% compression (redundancy) vs. complexity -- appear to balance each other, butnot necessarily.

@@

The above prior studies are consistent with two potential descriptions of how
cognitive processes are reflects in brain activity patterns. One possibility is
that the information rate of brain activity increases during more complex or
higher-level cognitive processing. If so, then the ability to reliably decode
cognitive states from brain activity patterns should improve with task
complexity or with the level (or ``depth'') of cognitive processing. A second
possibility is that the compressibility of brain activity patterns increases
during more complex or higher-level cognitive processing. If so, then
individual features of brain recordings, or compressed representations of brain
recordings, should carry more information during complex or high-level (versus
simple or low-level) cognitive tasks.

@@

@@

We used an across-participant correlation-based classifier to decode which
stimulus timepoint matched each timepoint's neural pattern. For a given value
of $k$ (i.e., number of principal components), we first used group PCA to
project the data from each condition into a $k$-dimensional space. Next, we
divided the participants into two groups: a template group,
$\mathcal{G}_{\mathrm{template}}$ (i.e., training data), and a to-be-decoded
group, $\mathcal{G}_{\mathrm{decode}}$ (i.e., test data). We averaged the
projected data within each group to obtain a single $T$ by $k$ matrix for each
group. Next, we correlated the rows of the two averaged matrices to form a $T$
by $T$ decoding matrix, $\mathbf{\Lambda}$. In this way, the rows of
$\mathbf{\Lambda}$ reflected timepoints from the template group, while the
columns reflected timepoints from the to-be-decoded group. We used
$\mathbf{\Lambda}$ to assign temporal labels to each timepoint (row) from the
test group's matrix, using the row of the training group's matrix with which it
was most highly correlated. We repeated this decoding procedure, but using
$\mathcal{G}_{\mathrm{decode}}$ as the template group and
$\mathcal{G}_{\mathrm{template}}$ as the to-be-decoded group. Given the true
timepoint labels (for each group), we defined the decoding accuracy as the
average proportion of correctly decoded timepoints, across both groups (where
chance perfomance is $\frac{1}{T}$). In Figures~\ref{fig:inflection}
and~\ref{fig:inflection-thirds} we report the decoding accuracy for each
condition and value of $k$, averaged across $n = 1,000$ cross validation folds.

>>>>

\documentclass[english, 11pt]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\usepackage{soul}
\linenumbers

@@

\bigskip
\noindent
\textbf{Keywords: information, compression, temporal decoding, dimensionality reduction, neuroimaging}

@@

@@

The above studies are consistent with two potential descriptions of how
cognitive processes are reflects in brain activity patterns. One possibility is
that the information rate of brain activity increases during more complex or
higher-level cognitive processing. If so, then the ability to reliably decode
cognitive states from brain activity patterns should improve with task
complexity or with the level (or ``depth'') of cognitive processing. A second
possibility is that the compressibility of brain activity patterns increases
during more complex or higher-level cognitive processing. If so, then
individual features of brain recordings, or compressed representations of brain
recordings, should carry more information during complex or high-level (versus
simple or low-level) cognitive tasks.

@@

We sought to understand whether higher-level cognition is reflected by more
reliable and informative brain activity patterns, and how compressibility of
brain activity patterns relates to cognitive complexity. We developed a
computational framework for systematically assessing the informativeness and
compressibility of brain activity patterns recorded under different cognitive
circumstances. We used across-participant decoding accuracy (see
\textit{Forward inference and decoding accuracy}) as a proxy for
informativeness. To estimate the compressibility of the brain patterns, we used
group principal components analysis (PCA) to project the brain patterns into
$k$-dimensional spaces, for different values of $k$ (see \textit{Hierarchical
Topographic Factor Analysis (HTFA)} and \textit{Principal components analysis
(PCA)}). For more compressible brain patterns, decoding accuracy should be more
robust to small values of $k$.

We analyzed a dataset collected by \cite{SimoEtal16} that comprised four
experimental conditions. These conditions exposed participants to stimuli that
systematically varied in cognitive engagement. In the \textit{intact}
experimental condition, participants listened to an audio recording of a
10-minute story. In the \textit{paragraph}-scrambled experimental condition,
partipants listened to a temporally scrambled version of the story, where the
paragraphs occurred out of order, but where the same set of paragraphs was
presented over the entire listening interval. All participants in this
condition experienced the scrambled paragraphs in the same order. In the
\textit{word}-scrambled experimental condition, participants listened to a
temporally scrambled version of the story, where the words occurred in a random
order. Again, all participants in this condition experienced the scrambled
words in the same order. Finally, in the \textit{rest} experimental condition,
participants lay in the scanner with no overt stimulus, while keeping their
eyes open and blinking as needed. This public dataset provided a convenient
means for testing our hypothesis that different levels of cognitive processing
and engagement affect how informative and compressible the associated brain
patterns are.

@@

We used an across-participant correlation-based classifier to decode which
stimulus timepoint matched each timepoint's neural pattern. For a given value
of $k$ (i.e., number of principal components), we first used group PCA to
project the data from each condition into a $k$-dimensional space. Next, we
divided the participants into two groups: a template group,
$\mathcal{G}_{\mathrm{template}}$ (i.e., training data), and a to-be-decoded
group, $\mathcal{G}_{\mathrm{decode}}$ (i.e., test data). We averaged the
projected data within each group to obtain a single $T$ by $k$ matrix for each
group. Next, we correlated the rows of the two averaged matrices to form a $T$
by $T$ decoding matrix, $\mathbf{\Lambda}$. In this way, the rows of
$\mathbf{\Lambda}$ reflected timepoints from the template group, while the
columns reflected timepoints from the to-be-decoded group. We used
$\mathbf{\Lambda}$ to assign temporal labels to each timepoint (row) from the
test group's matrix, using the row of the training group's matrix with which it
was most highly correlated. We repeated this decoding procedure, but using
$\mathcal{G}_{\mathrm{decode}}$ as the template group and
$\mathcal{G}_{\mathrm{template}}$ as the to-be-decoded group. Given the true
timepoint labels (for each group), we defined the decoding accuracy as the
average proportion of correctly decoded timepoints, across both groups (where
chance perfomance is $\frac{1}{T}$). In Figures~\ref{fig:inflection}
and~\ref{fig:inflection-thirds} we report the decoding accuracy for each
condition and value of $k$, averaged across $n = 100$ cross validation folds.
